---
title: "Final Report"
author: "Andrew"
date: "May 12, 2016"
output: html_document
---

```{r setup, echo = FALSE, eval=TRUE, include=FALSE}

suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(plotly))

```

**Introduction:**

The Spatafora lab conducted an RNA-Sequencing experiment to see the effect of 
knocking out the SloR gene on genetic expression. SloR is a metalloregulator 
which regulates the expression of certain genes in a manganese depedent fashion.
The goal was to determine which genes, or types of genes are regulated by SloR.
SloR knockouts of *Streptococcus mutans* were created and labeled "GMS584".
The wild-type (natural) organism was labeled "UA159". The goal is to determine
which genes or groups of genes are differentially expressed between these two
strains of *S.mutans*. In order to do this, four steps were taken. First, 
all the genes found in *S.mutans*, the counts matrix, and the genome
alignments were organized into clean tables. Secondly, edgeR, a package under
Bioconductor, was used to determine fold-change and significance in genetic 
expression for each gene between UA159 and GMS584. Thirdly, gene ontology data
was found from the Gene Ontology Consortium and differential expression for
biological functions were found. 



```{r, echo = FALSE, eval=TRUE, warning=FALSE}

#Cleaning up the Data:

#Genome annotation from NCBI Genome Database
annotation <- read.delim("data/annotation.txt", header=FALSE, comment.char="#")

annotation <- annotation %>%
  select(V6, V7, V8, V9, V12)

names(annotation)<- c("NCBI_ID", "Gene", "SMU", "A_num", "Desc")

annotation$NCBI_ID <- as.numeric(as.character(annotation$NCBI_ID))


#Mapping back to the Genome (done by Dr. Robert Haney)

descripts <- read.csv("data/gene_descriptions.csv")

gene_ID <- data.frame(do.call('rbind',
                              strsplit(as.character(
                                descripts$Description),';',fixed=TRUE)))

#Cleaning further to remove descripters
gene_ID$X1 <- gene_ID$X1 %>%
  str_replace_all("ID=gene", "") %>%
  as.numeric(as.character())
gene_ID$X3 <- gene_ID$X3 %>%
  str_replace_all("Dbxref=GeneID:", "") %>%
  as.numeric(as.character())

names(gene_ID)[0:5] <- c("gene_num", "common", "NCBI_ID", "rando", "SMU")

dbx_ID <- gene_ID %>%
  select(gene_num, NCBI_ID)

#Selective only for the rRNAs and tRNAs
dbx_RNA <- gene_ID %>%
  select(NCBI_ID, SMU) %>%
  filter(NCBI_ID >2000000)

#Remove the locus_tag bit before SMU
dbx_RNA$SMU <- dbx_RNA$SMU %>%
  str_replace_all("locus_tag=", "")

#bind NCBI's annotations with the rRNA and tRNA annotations from R.Haney
annotation <- bind_rows(dbx_RNA, annotation)

annotation2 <- left_join(dbx_ID, annotation, by ="NCBI_ID")


#------
#Analyzing the count matrix

#There are four replicates for each strain. Nothing to do with Mn.
counts <- read.delim("data/Smutans_UA159_GMS584.genes.counts.matrix",
                     row.names="Symbol")

#Building DGEList.
Strand <- factor(c("UA159", "UA159", "UA159", "UA159", "GMS584"
                   ,"GMS584","GMS584","GMS584"))

y<- DGEList(counts,group=Strand)


#Filter transcripts with very low expression counts >1CPM in at least 4 replicates
keep <- rowSums(cpm(y)>1) >= 4
y <- y[keep,]

#recompute library sized
y$samples$lib.size <- colSums(y$counts)

#Normalizes for RNA composition by library sizes 
y <- calcNormFactors(y)

#Dispersion means biological coeffient of variation (BCV) squared.

y <- estimateCommonDisp(y, verbose = TRUE)
y <- estimateTrendedDisp(y)
y<- estimateTagwiseDisp(y)


#Plot estimated dispersions
plotBCV(y, log = 'y')

```

\n
The produced plot above shows a graph of biological variation between samples.
Biological variation separates itself from any technical variation. It 
measures any variation in the data between replicates. You want this number to
be low, since it will eliminate any differential expression that results from
variations between replicates. 

The dispersion of a gene is simply another measure of a gene's variance and it 
is used by DESeq to model the overall variance of a gene's count values. The 
dispersion can be interpreted as the square of the coefficient of biological 
variation (e.g. the difference in counts between two biological replicates 
is 15.5%% so the gene's dispersion is (0.1558)^2 = 0.0246). Comparing this to
published data for prokayotes, this number is relatively low, which signifies
that the data is consistent and that variation in replicates is minimal. 



```{r, echo = FALSE, eval=TRUE, warning=FALSE}

#Compute exact gene wise tests for differential expression amount treatments
et <- exactTest(y, pair = c("UA159", "GMS584"))
#IMPORTANT:  Note that the first group listed in the pair is the baseline for the comparison—so c("UA159","GMS584") then is the comparison of GMS584 to UA159, so genes with positive log-fold change are up-regulated in group GMS584 compared with group UA159 (and vice versa for genes with negative log-fold change).

topTags(et)

```

Above you can see the top 10 genes in terms of signficance based off their 
P values. Gene numbers are shown on the left. This numbers provide no 
information on the actual gene and are merely place holders until the genes can
be mapped to the reference genome. Second to the P value, logFC is an important
value. This is the log-fold change between GMS584 to UA159. In other words, a 
logFC of -6.4 means that gene609 in UA159 is 6.4 times more abundant than it is 
in GMS584. logCPM is the average log counts per million of that gene for
UA159 and GMS584 and FDR is a measure of confidence. It measures for any 
variation that could be due to chance. These are all low FDR values. 



```{r, echo = FALSE, eval=TRUE, warning=FALSE}

#Summarize DE genes at 5% FCR (default is 0.05)
summary(dt <- decideTestsDGE(et))

#Gives the names of DE genes and intergenic
isDE <- as.logical(dt)
DEnames <- rownames(y)[isDE]

#Plots each transcript and it's log FC. Shows DE genes in red
plotSmear(et, de.tags = DEnames)
abline(h=c(-1,1), col="blue")

#Organizing tTags output
tTags <- topTags(et,n=NULL)

write.table(tTags[tTags$table$FDR <= 0.05,], file='tTags', sep='\t', quote=F, row.names=T)

tTags <- read.delim("tTags", row.names=1)

tTags <- mutate(tTags, gene_num = rownames(tTags))
tTags$gene_num <- tTags$gene %>%
  str_replace_all("gene", "") %>%
  as.numeric(as.character())

tTags <- tTags %>% 
  filter(!is.na(gene_num))

#Bringing genome alignments to the tTags data 
tTags_annot <- left_join(tTags, annotation2, by = "gene_num")
```


The plot above shows the log fold change graphed against log counts-per-million.
This shows the differential expression of each gene that was read by the 
Illumina Hi-Seq. In other words this includes all the genes and intergenic 
regions. Intergenic regions includes the non-coding parts of the transcriptone 
meaning these do not develop into proteins, but rather are functional in their
RNA form. These include small RNAs, ribosomal RNA, and tRNA. Differentially 
expressed (DE) genes are in red, which is based off a p-value of 0.005. The blue 
lines are there to show DE genes that have a logFC of greater than one. As you
can see from this graph there are about equal amounts of overexpressed and 
underexpressed genes in UA159 compared to GMS584. logFC less than 0 correspond
to an upregulation in UA159. logFC greater than 0 correspond to upregulation
in GMS584. 

Lastly, the summary function shows the number of genes that are overexpressed
in UA159 compared to the number of genes that are downregulated compared to the
number of genes that are not differentially expressed. As you can see the number
of upregulated and downregulated genes are almost equal. 


```{r, echo = FALSE, eval=TRUE, warning=FALSE}

#------------CPM for graphical comparison

logcpm <- cpm(y, prior.count=2, log=TRUE)
rn_cpm <- as.character(rownames(logcpm))

logcpm <- logcpm %>% 
  data.frame() %>%
  tbl_df()

logcpm$gene_num <- as.character(rn_cpm) %>% 
  str_replace_all("gene", "") %>%
  as.numeric(as.character())

norm_counts <- logcpm

norm_counts <- norm_counts %>% 
  mutate(avgCPM_UA159 = (UA159.1 + UA159.2 + UA159.3 + UA159.4)/4) %>% 
  mutate(avgCPM_GMS584 = (GMS584.1 + GMS584.2 + GMS584.3 + GMS584.4)/4) %>% 
  filter(!is.na(gene_num))
  
norm_counts <- norm_counts %>% 
  select(gene_num, avgCPM_UA159, avgCPM_GMS584)

#FINAL BEAUTIFUL FINISHED TABLE
tTags_annot <- 
  left_join(tTags_annot, norm_counts, by = "gene_num") %>% 
  select(SMU, Gene, NCBI_ID, logFC, logCPM, PValue, 
         FDR, Desc, avgCPM_UA159, avgCPM_GMS584) %>% 
  arrange(SMU)

#To be used in Shiny App
write.csv(file="data/tTags_annot.csv", tTags_annot, row.names = FALSE)


#Looking at overall trends in the data with CPM UA159 vs. CPM GMS584
t <- ggplot(data= tTags_annot, aes(x = avgCPM_UA159, 
                                   y = avgCPM_GMS584, color = logFC))  


t + geom_hex() + 
  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0) +
  labs(title = "CPM Between UA159 and GMS584 with FC",
       x= "UA159(CPM)", y = "GMS584(CPM)")

```

This plot is very similar to the plot from above, but it consolidates much of
the information into a more approachable image. This emphasisizes the areas 
of the plot that have the highest concentration of genes. Most of the genes are 
found to staddle the line of y = x, which means that the most of the genes are 
found to have the same expression in UA159 and GMS584. In addition, you can see
that most of the genes have average CPM betwen 5 and 7. Lastly, you can see 
that the number of genes that are overexpressed in UA159 are more differentailly
expressed than the genes that are overexpressed in GMS584; in other words, the 
genes that are overexpressed in UA159 have a higher average logFC than the 
genes that are overexpressed in GMS584. 



```{r, echo = FALSE, eval=TRUE, warning=FALSE}
#--------Linking GENE ONTOLOGY

#Downloaded from Gene Ontology Consortium
obo <- readLines("go.obo.txt")

id_lines <- which(str_detect(obo, "id: GO:") & !str_detect(obo, "alt_id: GO:"))
name_lines <- id_lines + 1


GO_name <-
  data_frame(
    GO=obo[id_lines],
    name=obo[name_lines]
)

GO_name$GO <- GO_name$GO %>% 
  str_replace_all("id: ", "")

GO_name$name <- GO_name$name %>% 
  str_replace_all("name: ", "")


#bringing SMU to GO terms to Biological function

GO_terms <- read.delim("data/GO_terms.txt", header=FALSE) 

names(GO_terms)[1:2] <- c("SMU", "GO")

GO_terms <- left_join(GO_terms, GO_name, by = "GO")

GO_SMU <- right_join(GO_terms, tTags_annot, by= "SMU")

GO_n <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc) %>% 
  group_by(name) %>% 
  tally() %>% 
  top_n(20)

GO_PValue <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc) %>% 
  group_by(name) %>% 
  summarise(avg_p = mean(PValue))

#Get the most common and most significant pvalues
GO_n <- right_join(GO_PValue, GO_n, by = "name") %>% 
  arrange(avg_p) %>% 
  head(9)

to_graph <- GO_SMU %>% 
  select(SMU, GO, logFC, name, Desc, avgCPM_UA159, avgCPM_GMS584) %>% 
  filter(name %in%GO_n$name)

ggplot(to_graph, aes(x = avgCPM_UA159, y = avgCPM_GMS584, color = logFC)) + 
  geom_point() +
  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0,20)) +
  labs(title = "Differential Expression of GMS584 vs. UA159 Based off 
       Most Common GO terms and Highest P values",x = "UA159 CPM", y = "GMS584 CPM") +
  facet_wrap(~name)

#Most significant but not abundant

GO_n2 <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc) %>% 
  group_by(name) %>% 
  tally() %>% 
  filter(n>5)

GO_PValue2 <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc) %>% 
  group_by(name) %>% 
  summarise(avg_p = mean(PValue))

GO_n2 <- right_join(GO_PValue, GO_n2, by = "name") %>% 
  arrange(avg_p) %>% 
  head(9)

to_graph2 <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc, avgCPM_UA159, avgCPM_GMS584) %>% 
  filter(name %in%GO_n2$name)


ggplot(to_graph2, aes(x = avgCPM_UA159, y = avgCPM_GMS584, color = logFC)) + 
  geom_point() +
  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0,20)) +
  labs(title = "Differential Expression of GMS584 vs.UA159 by 
       P Values of GO terms Greater than 5", x = "UA159 CPM", y = "GMS584 CPM") +
  facet_wrap(~name)



#this didn't come up because there is no inclusive GO term for genetic competence
gene_comp <- GO_SMU %>% 
  select(SMU, logFC, logCPM, PValue, GO, name, Desc, avgCPM_UA159, avgCPM_GMS584) %>% 
  filter(str_detect(Desc,"competence"))

ggplot(gene_comp, aes(x = avgCPM_UA159, y = avgCPM_GMS584, color = logFC)) + 
  geom_point() +
  geom_abline(mapping = NULL, data = NULL, slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0,20)) +
  scale_colour_gradient2(low="#520101", high="blue") +
  labs(title = "Genetic Competence", x = "UA159 CPM", y = "GMS584 CPM")
#WOW! ALL genes related to competence are massively downregulated in GMS584

#To be used in Shiny App
write.csv(file="data/GO_SMU.csv", GO_SMU, row.names = FALSE)

```

These three plots are used to show trends based of Gene ontology terms. Gene 
Ontology cover three main domains, biological process, molecular function,
and cellular component. It organizes genes into functional terms which can be 
used to observe large themes in differential expression of like genes with 
similar functions.

The first plot shows the most common GO terms with the top 9 p-values and the
genes that are differentially expressed. Genes are that below the line of y=x
are upregulated in UA159 and downregulated in GMS584. Genes that are above the
line y=x are downregualted in UA159 and upregulated in GMS584. This tells us a
big conclusion- any upregulation in GMS584 shows us a gene that is normally
under strict regulation by SloR (the regulator that was knocked out in this 
model). If the gene is upregulated in UA159, then the gene in directly or
indirectly initiated by SloR. From this faceted plot, you can see that the 
genes involved with phosphotransferase activity are downregulated in GMS584,
which means that these genes could be positively regulated, meaning that they 
could be initiated by the presence of SloR. This is the opposite for transporter
activity, which means SloR negatively regulates these genes, meaning SloR 
blocks transcription of these genes. 

The next plot shows the differential expression of GMS584 vs UA159 by GO terms
with the most significant (lowest) p-values that have at least 5 genes. This
plot shows similar results for the genes involved in phosphotransferase activity 
and transporter activity. Interestingly, DNA repair and defense response to 
bacterium seem to be upregulated in UA159 meaning that these genes are under
SloR activation, meaning SloR directly or indirectly activates their 
transcription. 

Lastly, I wanted to look at genetic competence, but this highlighted some of
the problems when working with gene ontology. The terms are not complete; for
genetic competence there is no concise term to use. This plot uses gene 
annotations that are included with the **S.mutans** genome from the NCBI database. 
This plot shows the genes that include the term "competence" in their 
description. As you can see the genes involved in genetic competence are 
massively downregulated in GMS584, which highlights the importance of SloR in
bringing a bacterium to competence. SloR obviously some how activates the 
expression of genes involved in making a bacterium competent. Competence is 
a point in the growth of a bacterium when it is able to absorb foreign DNA
and incorporate it into its own transcription machinery. 

**Concluding Remarks:**

This investigation into the transcriptome of *Streptococcus mutans* proved
very valuable. New differentially expressed genes have been identified such 
as those involved in defense response to bacterium and DNA repair. The 
connection between these genes and SloR were not known before. These new 
findings will lead to further experiments such as qRT-PCR which will be able
to experimentally confirm these findings. In addition, the RNA-Seq data
confirmed previous hypotheses and RT-PCR experiments for genes such as *spx*,
and *comC*. 






