---
title: "HOCR Slides"
author: "Phil Hoxie"
date: "May 16, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, results='hide', message=FALSE, error=FALSE}
## Packages
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(leaflet)
library(ggmap)
library(maps)
library(maptools)
library(plotly)
library(rgdal)
library(rgeos)
library(scales)
library(rvest)
library(stringr)
library(knitr)
## Data

athletes <- read.csv(file = "data/athletes_hocr.csv", header = TRUE)
clubs <- read.csv(file = "data/clubs_hocr.csv", header = TRUE)
events <- read.csv(file = "data/events_hocr.csv", header = TRUE)
races <- read.csv(file = "data/races_hocr.csv", header = TRUE)
results <- read.csv(file = "data/results.entry.csv", header = TRUE, stringsAsFactors = FALSE)

race.details <- left_join(events, races, by = "event_id")

markers <- read.csv("data/markers.csv", header = TRUE, stringsAsFactors = FALSE)

penalty.markers <- read.csv("data/penalty.markers.csv", header = TRUE, stringsAsFactors = FALSE)

## results by race

## Men
m.alum.8 <- read.csv(file = "data/m.alum.8.csv", header = TRUE)
m.chmp.4 <- read.csv(file = "data/m.chmp.4.csv", header = TRUE)
m.chmp.8 <- read.csv(file = "data/m.chmp.8.csv", header = TRUE)
m.club.4 <- read.csv(file = "data/m.chmp.4.csv", header = TRUE)
m.club.8 <- read.csv(file = "data/m.chmp.8.csv", header = TRUE)
m.coll.4 <- read.csv(file = "data/m.college.4.csv", header = TRUE)
m.coll.8 <- read.csv(file = "data/m.college.8.csv", header = TRUE)
m.lght.4 <- read.csv(file = "data/m.light.4.csv", header = TRUE)
m.lght.8 <- read.csv(file = "data/m.light.8.csv", header = TRUE)

## Women

w.alum.8 <- read.csv(file = "data/w.alum.8.csv", header = TRUE)
w.chmp.4 <- read.csv(file = "data/w.chmp.4.csv", header = TRUE)
w.chmp.8 <- read.csv(file = "data/w.chmp.8.csv", header = TRUE)
w.club.4 <- read.csv(file = "data/w.chmp.4.csv", header = TRUE)
w.club.8 <- read.csv(file = "data/w.chmp.8.csv", header = TRUE)
w.coll.4 <- read.csv(file = "data/w.college.4.csv", header = TRUE)
w.coll.8 <- read.csv(file = "data/w.college.8.csv", header = TRUE)
w.lght.4 <- read.csv(file = "data/w.light.4.csv", header = TRUE)
w.lght.8 <- read.csv(file = "data/w.light.8.csv", header = TRUE)

## Master results sheet

## First we must delete the extra columns and standardize the column names
names(m.alum.8)
m.alum.8 <- m.alum.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, split4, cum4)

names(w.alum.8)
w.alum.8 <- w.alum.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         X.18, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)

w.alum.8 <- w.alum.8 %>% 
  rename(split1 = X.18, split4 = splitF)

names(m.chmp.4)
m.chmp.4 <- m.chmp.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.chmp.4 <- m.chmp.4 %>% 
  rename(split4 = splitF)

names(w.chmp.4)
w.chmp.4 <- w.chmp.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.chmp.4 <- w.chmp.4 %>% 
  rename(split4 = splitF)

names(m.chmp.8)
m.chmp.8 <- m.chmp.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.chmp.8 <- m.chmp.8 %>% 
  rename(split4 = splitF)

names(w.chmp.8)
w.chmp.8 <- w.chmp.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.chmp.8 <- w.chmp.8 %>% 
  rename(split4 = splitF)

names(m.club.4)
m.club.4 <- m.club.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.club.4 <- m.club.4 %>% 
  rename(split4 = splitF)

names(w.club.4)
w.club.4 <- w.club.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.club.4 <- w.club.4 %>% 
  rename(split4 = splitF)

names(m.club.8)
m.club.8 <- m.club.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.club.8 <- m.club.8 %>% 
  rename(split4 = splitF)

names(w.club.8)
w.club.8 <- w.club.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.club.8 <- w.club.8 %>% 
  rename(split4 = splitF)

names(m.coll.4)
m.coll.4 <- m.coll.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.coll.4 <- m.coll.4 %>% 
  rename(split4 = splitF)

names(w.coll.4)
w.coll.4 <- w.coll.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.coll.4 <- w.coll.4 %>% 
  rename(split4 = splitF)

m.coll.8 <- m.coll.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.coll.8 <- m.coll.8 %>% 
  rename(split4 = splitF)

w.coll.8 <- w.coll.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.coll.8 <- w.coll.8 %>% 
  rename(split4 = splitF)

m.lght.4 <- m.lght.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.lght.4 <- m.lght.4 %>% 
  rename(split4 = splitF)

w.lght.4 <- w.lght.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.lght.4 <- w.lght.4 %>% 
  rename(split4 = splitF)

m.lght.8 <- m.lght.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.lght.8 <- m.lght.8 %>% 
  rename(split4 = splitF)

w.lght.8 <- w.lght.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.lght.8 <- w.lght.8 %>% 
  rename(split4 = splitF)

## let's make the master results sheet

regatta <- bind_rows(list(m.alum.8, w.alum.8, 
                          m.chmp.4, w.chmp.4, 
                          m.chmp.8, w.chmp.8,
                          m.club.4, w.club.4,
                          m.club.8, w.club.8,
                          m.coll.4, w.coll.4,
                          m.coll.8, w.coll.8,
                          m.lght.4, w.lght.4,
                          m.lght.8, w.lght.8))

regatta <- regatta %>% 
  select(event_id, place, status, 
         lane_bow, club, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, split4, cum4)

## Excel created some extra rows, lets drop them
regatta <- regatta %>% 
  mutate(event_id = as.character(event_id))
regatta <- regatta %>%  
  filter(!is.na(event_id))

## here is the final sheet
regatta.full <- left_join(regatta, race.details, by = "event_id")

## athletes

## US cities from http://simplemaps.com/resources/us-cities-data 
## lat long from http://www.latlong.net/

cities <- read.csv(file = "data/cities.csv", head=TRUE, stringsAsFactors = FALSE)

## state abbreviations from: http://www.infoplease.com/ipa/A0110468.html

abbrev <- read.csv(file = "data/state.abbreviations.csv", header = TRUE, stringsAsFactors = FALSE)

clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}

cities <- cities %>% 
  mutate(state.clean = clean_text(state), 
         city.clean = clean_text(city))

cities.tidy <- cities %>% 
  mutate(full.name = paste(city.clean,state.clean, sep = ", "))

cities.tidy <- cities.tidy %>% 
  distinct(full.name)

athletes <- athletes %>% 
  mutate(state.clean = clean_text(state.province), 
         city.clean = clean_text(city))

athletes.us <- athletes %>% 
  filter(country=="US") %>% 
  group_by(state.clean, city.clean, city, state.province) %>% 
  tally()
## drop 75 athletes who did not report their city
athletes.us <- athletes.us %>% 
  filter(city!="")

## state abbreviations

abbrev <- abbrev %>% 
  mutate(state.clean = clean_text(State), 
         code = clean_text(Postal.Code))

athletes.us <- left_join(athletes.us, abbrev, by=c("state.clean"))


athletes <- athletes %>% 
  mutate(club = as.character(club), 
         city = as.character(city), 
         state.province = as.character(state.province),
         country = as.character(country))

clubs.affiliated <- clubs %>% 
  filter(club != "Unaffiliated (usa)")

clubs.by.state <- clubs.affiliated %>% 
  filter(federation=="USA") %>% 
  group_by(state.province) %>% 
  tally() %>% 
  arrange(desc(n))
clubs.by.state <- clubs.by.state %>% 
  rename(clubs = n)

athletes.by.state <- athletes %>% 
  filter(country=="US") %>% 
  group_by(state.province) %>% 
  tally() %>% 
  arrange(desc(n))

athletes.by.state <- athletes.by.state %>% 
  rename(athletes = n)


participation <- left_join(clubs.by.state, athletes.by.state, by = "state.province")
participation <- participation %>% 
  mutate(hover = paste("clubs: ", clubs, "<br>",
                       "athletes: ", athletes))
participation <- participation %>% 
  mutate(state.clean = clean_text(state.province))
participation <- left_join(participation, abbrev, by = c("state.clean"="state.clean"))

athletes.cities <- left_join(athletes.us, cities.tidy, 
                             by = c("code" = "state.clean", "city.clean" = "city.clean"))

```

## Crash Course in Rowing Terms

Boat Types:
Coxed
-4+
-8+
Uncoxed
-Single
-Double
-Quad

The 500m split
-shown in erg output
-1:50 means "You are rowing 500 meters every 1 min and 50 seconds"
-standardizes the concept of average work
  

## The Head of the Charles

-The largest regatta in the world
-10,234 Rowers
-25 Countries
-781 Clubs
-4800 meter "Head" Race
-boats are staggered by seed position 
-Serpantine race course
-Bridges cuase bottlenecking 
  
## What Countries are represented? 

```{R, echo = FALSE}
clubs.by.country <- clubs %>% 
  group_by(country) %>% 
  tally() %>% 
  arrange(desc(n))

kable(head(clubs.by.country), col.names = c("coutry", "Number of Clubs"))
```
## What Countreis are represented? 

```{R, echo = FALSE}
c.b.country <- 
  ggplot(data = clubs.by.country, aes(x=factor(country, levels=unique(country)), y=n))+
  geom_bar(stat="identity")+
  ggtitle("HOCR51 Rowing Clubs by Country")+
  ylab("Number of Clubs")+
  xlab("Country")
c.b.country

```

## What Countries are the Competitors from? 

```{R, echo = FALSE}
athletes.by.country <- athletes %>% 
  group_by(country) %>% 
  tally() %>% 
  arrange(desc(n))
kable(head(athletes.by.country), col.names = c("coutry", "Number of Rowers"))
```
## What countries are the Competitors from? 

```{R, echo = FALSE}

r.b.country <- 
  ggplot(data = athletes.by.country, aes(x=factor(country, levels=unique(country)), y=n))+
  geom_bar(stat="identity")+
  ggtitle("HOCR51 Rowers by Country")+
  ylab("Number of Clubs")
r.b.country
```

## What States are Represented? 

```{R, echo = FALSE}

map.clubs <- plot_ly(data = participation, z = athletes, text = hover, locations = Postal.Code, type = 'choropleth',
        locationmode = 'USA-states', color = athletes, colors = 'Purples',
        marker = list(line = list(color = toRGB("black"), width = 1)), colorbar = list(title = "Number of Rowers")) %>%
  layout(title = '2015 HOCR Participation by State<br>(Hover for breakdown)', geo = list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = TRUE,
    lakecolor = toRGB('white')
  ))
map.clubs
```
## What Cities are Represented? 

```{R, echo = FALSE, fig_height = 6}

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray85"),
  subunitwidth = 1,
  countrywidth = 1,
  subunitcolor = toRGB("white"),
  countrycolor = toRGB("white")
)

map.all <- plot_ly(athletes.cities, lon = lng, lat = lat, text = paste(full.name, "<br>", n),
        marker = list(size = n/10),
        type = 'scattergeo', locationmode = 'USA-states') %>%
  layout(title = '2015 Head of the Charles Rowers', geo = g)
map.all
```

## Research Questions

From the persepctive of a coach:

  1. What races are the most competitive?
  2. How Should I tell my Coxswain to Row the Course? 
  3. How Should I tell my rowers to row the course? 
