---
title: "HOCR Slides"
author: "Phil Hoxie"
date: "May 16, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, results='hide', message=FALSE, error=FALSE, warning=FALSE}
## Packages
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(leaflet)
library(ggmap)
library(maps)
library(maptools)
library(plotly)
library(rgdal)
library(rgeos)
library(scales)
library(rvest)
library(stringr)
library(knitr)
## Data
split.polygons <- read.csv("data/split.polygons.csv", header = TRUE, stringsAsFactors = FALSE)

athletes <- read.csv(file = "data/athletes_hocr.csv", header = TRUE)
clubs <- read.csv(file = "data/clubs_hocr.csv", header = TRUE)
events <- read.csv(file = "data/events_hocr.csv", header = TRUE)
races <- read.csv(file = "data/races_hocr.csv", header = TRUE)
results <- read.csv(file = "data/results.entry.csv", header = TRUE, stringsAsFactors = FALSE)

race.details <- left_join(events, races, by = "event_id")

markers <- read.csv("data/markers.csv", header = TRUE, stringsAsFactors = FALSE)

penalty.markers <- read.csv("data/penalty.markers.csv", header = TRUE, stringsAsFactors = FALSE)

## results by race

## Men
m.alum.8 <- read.csv(file = "data/m.alum.8.csv", header = TRUE)
m.chmp.4 <- read.csv(file = "data/m.chmp.4.csv", header = TRUE)
m.chmp.8 <- read.csv(file = "data/m.chmp.8.csv", header = TRUE)
m.club.4 <- read.csv(file = "data/m.chmp.4.csv", header = TRUE)
m.club.8 <- read.csv(file = "data/m.chmp.8.csv", header = TRUE)
m.coll.4 <- read.csv(file = "data/m.college.4.csv", header = TRUE)
m.coll.8 <- read.csv(file = "data/m.college.8.csv", header = TRUE)
m.lght.4 <- read.csv(file = "data/m.light.4.csv", header = TRUE)
m.lght.8 <- read.csv(file = "data/m.light.8.csv", header = TRUE)

## Women

w.alum.8 <- read.csv(file = "data/w.alum.8.csv", header = TRUE)
w.chmp.4 <- read.csv(file = "data/w.chmp.4.csv", header = TRUE)
w.chmp.8 <- read.csv(file = "data/w.chmp.8.csv", header = TRUE)
w.club.4 <- read.csv(file = "data/w.chmp.4.csv", header = TRUE)
w.club.8 <- read.csv(file = "data/w.chmp.8.csv", header = TRUE)
w.coll.4 <- read.csv(file = "data/w.college.4.csv", header = TRUE)
w.coll.8 <- read.csv(file = "data/w.college.8.csv", header = TRUE)
w.lght.4 <- read.csv(file = "data/w.light.4.csv", header = TRUE)
w.lght.8 <- read.csv(file = "data/w.light.8.csv", header = TRUE)

## Master results sheet

## First we must delete the extra columns and standardize the column names
names(m.alum.8)
m.alum.8 <- m.alum.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, split4, cum4)

names(w.alum.8)
w.alum.8 <- w.alum.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         X.18, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)

w.alum.8 <- w.alum.8 %>% 
  rename(split1 = X.18, split4 = splitF)

names(m.chmp.4)
m.chmp.4 <- m.chmp.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.chmp.4 <- m.chmp.4 %>% 
  rename(split4 = splitF)

names(w.chmp.4)
w.chmp.4 <- w.chmp.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.chmp.4 <- w.chmp.4 %>% 
  rename(split4 = splitF)

names(m.chmp.8)
m.chmp.8 <- m.chmp.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.chmp.8 <- m.chmp.8 %>% 
  rename(split4 = splitF)

names(w.chmp.8)
w.chmp.8 <- w.chmp.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.chmp.8 <- w.chmp.8 %>% 
  rename(split4 = splitF)

names(m.club.4)
m.club.4 <- m.club.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.club.4 <- m.club.4 %>% 
  rename(split4 = splitF)

names(w.club.4)
w.club.4 <- w.club.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.club.4 <- w.club.4 %>% 
  rename(split4 = splitF)

names(m.club.8)
m.club.8 <- m.club.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.club.8 <- m.club.8 %>% 
  rename(split4 = splitF)

names(w.club.8)
w.club.8 <- w.club.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.club.8 <- w.club.8 %>% 
  rename(split4 = splitF)

names(m.coll.4)
m.coll.4 <- m.coll.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.coll.4 <- m.coll.4 %>% 
  rename(split4 = splitF)

names(w.coll.4)
w.coll.4 <- w.coll.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.coll.4 <- w.coll.4 %>% 
  rename(split4 = splitF)

m.coll.8 <- m.coll.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.coll.8 <- m.coll.8 %>% 
  rename(split4 = splitF)

w.coll.8 <- w.coll.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.coll.8 <- w.coll.8 %>% 
  rename(split4 = splitF)

m.lght.4 <- m.lght.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.lght.4 <- m.lght.4 %>% 
  rename(split4 = splitF)

w.lght.4 <- w.lght.4 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.lght.4 <- w.lght.4 %>% 
  rename(split4 = splitF)

m.lght.8 <- m.lght.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
m.lght.8 <- m.lght.8 %>% 
  rename(split4 = splitF)

w.lght.8 <- w.lght.8 %>% 
  select(event_id, place, status, 
         lane_bow, club, crew, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, splitF, cum4)
w.lght.8 <- w.lght.8 %>% 
  rename(split4 = splitF)

## let's make the master results sheet

regatta <- bind_rows(list(m.alum.8, w.alum.8, 
                          m.chmp.4, w.chmp.4, 
                          m.chmp.8, w.chmp.8,
                          m.club.4, w.club.4,
                          m.club.8, w.club.8,
                          m.coll.4, w.coll.4,
                          m.coll.8, w.coll.8,
                          m.lght.4, w.lght.4,
                          m.lght.8, w.lght.8))

regatta <- regatta %>% 
  select(event_id, place, status, 
         lane_bow, club, 
         rower, penalty, penalty_desc, 
         time_official, delta, percent, 
         split1, cum1, split2, cum2, 
         split3, cum3, split4, cum4)

## Excel created some extra rows, lets drop them
regatta <- regatta %>% 
  mutate(event_id = as.character(event_id))
regatta <- regatta %>%  
  filter(!is.na(event_id))

## here is the final sheet
regatta.full <- left_join(regatta, race.details, by = "event_id")

## athletes

## US cities from http://simplemaps.com/resources/us-cities-data 
## lat long from http://www.latlong.net/

cities <- read.csv(file = "data/cities.csv", head=TRUE, stringsAsFactors = FALSE)

## state abbreviations from: http://www.infoplease.com/ipa/A0110468.html

abbrev <- read.csv(file = "data/state.abbreviations.csv", header = TRUE, stringsAsFactors = FALSE)

clean_text <- function(text){
  text <- gsub("[^[:alnum:]]", "", text)
  text <- gsub(" ", "", text)
  text <- tolower(text)
  return(text)
}

cities <- cities %>% 
  mutate(state.clean = clean_text(state), 
         city.clean = clean_text(city))

cities.tidy <- cities %>% 
  mutate(full.name = paste(city.clean,state.clean, sep = ", "))

cities.tidy <- cities.tidy %>% 
  distinct(full.name)

athletes <- athletes %>% 
  mutate(state.clean = clean_text(state.province), 
         city.clean = clean_text(city))

athletes.us <- athletes %>% 
  filter(country=="US") %>% 
  group_by(state.clean, city.clean, city, state.province) %>% 
  tally()
## drop 75 athletes who did not report their city
athletes.us <- athletes.us %>% 
  filter(city!="")

## state abbreviations

abbrev <- abbrev %>% 
  mutate(state.clean = clean_text(State), 
         code = clean_text(Postal.Code))

athletes.us <- left_join(athletes.us, abbrev, by=c("state.clean"))


athletes <- athletes %>% 
  mutate(club = as.character(club), 
         city = as.character(city), 
         state.province = as.character(state.province),
         country = as.character(country))

clubs.affiliated <- clubs %>% 
  filter(club != "Unaffiliated (usa)")

clubs.by.state <- clubs.affiliated %>% 
  filter(federation=="USA") %>% 
  group_by(state.province) %>% 
  tally() %>% 
  arrange(desc(n))
clubs.by.state <- clubs.by.state %>% 
  rename(clubs = n)

athletes.by.state <- athletes %>% 
  filter(country=="US") %>% 
  group_by(state.province) %>% 
  tally() %>% 
  arrange(desc(n))

athletes.by.state <- athletes.by.state %>% 
  rename(athletes = n)


participation <- left_join(clubs.by.state, athletes.by.state, by = "state.province")
participation <- participation %>% 
  mutate(hover = paste("clubs: ", clubs, "<br>",
                       "athletes: ", athletes))
participation <- participation %>% 
  mutate(state.clean = clean_text(state.province))
participation <- left_join(participation, abbrev, by = c("state.clean"="state.clean"))

athletes.cities <- left_join(athletes.us, cities.tidy, 
                             by = c("code" = "state.clean", "city.clean" = "city.clean"))

athletes.by.country <- athletes %>% 
  group_by(country) %>% 
  tally() %>% 
  arrange(desc(n))

clubs.by.country <- clubs %>% 
  group_by(country) %>% 
  tally() %>% 
  arrange(desc(n))

```

## Crash Course in Rowing Terms

Coxed Boats

  - 4+
  - 8+

Uncoxed Boats
  
  - Single
  - Double
  - Quad

## Splits

The 500m split

  - shown in erg output
  - 1:50 means "You are rowing 500 meters every 1 min and 50 seconds"
  - standardizes the concept of average work
  

## The Head of the Charles

The largest regatta in the world

  + 10,234 Rowers
  + 25 Countries
  + 781 Clubs
  
4800 meter "Head" Race

  + boats are staggered by seed position 
  + Serpantine race course
  + Bridges cuase bottlenecking 


## What countries are the Competitors from? 

```{R, echo = FALSE}

r.b.country <- 
  ggplot(data = athletes.by.country, aes(x=factor(country, levels=unique(country)), y=n))+
  geom_bar(stat="identity")+
  ggtitle("HOCR51 Rowers by Country")+
  ylab("Number of Rowers")+
  xlab("Country")
r.b.country
```


## What States are Represented? 

```{R, echo = FALSE}

map.clubs <- plot_ly(data = participation, z = athletes, text = hover, locations = Postal.Code, type = 'choropleth',
        locationmode = 'USA-states', color = athletes, colors = 'Purples',
        marker = list(line = list(color = toRGB("black"), width = 1)), colorbar = list(title = "Number of Rowers")) %>%
  layout(title = '2015 HOCR Participation by State<br>(Hover for breakdown)', geo = list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = TRUE,
    lakecolor = toRGB('white')
  ))
map.clubs
```


## What Cities are Represented? 

```{R, echo = FALSE, fig_height = 6}

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray85"),
  subunitwidth = 1,
  countrywidth = 1,
  subunitcolor = toRGB("white"),
  countrycolor = toRGB("white")
)

map.all <- plot_ly(athletes.cities, lon = lng, lat = lat, text = paste(full.name, "<br>", n),
        marker = list(size = n/10),
        type = 'scattergeo', locationmode = 'USA-states') %>%
  layout(title = '2015 Head of the Charles Rowers', geo = g)
map.all
```


## Research Questions

From the persepctive of a coach:

  1. What races are the most competitive?
  2. How Should I tell my Coxswain to Row the Course? 
  3. How Should I tell my rowers to row the course? 
  
## The Coefficient of Variation

$$CV = \frac{\sigma}{\mu}$$

* CV = Stnd.dev / Average
* "standardizes" the standard deviation
* Good measure of relative dispersion/ congestion 


## 1. What Races are the most competitive

```{r, echo = FALSE, results = "hide"}
regatta.finishers <- regatta.full %>% 
  filter(status=="Finished")

##manipulate master sheet times

regatta.master.times <- regatta.finishers %>% 
  mutate(time_official = as.character(time_official)) %>% 
  separate(time_official,into = c("O.min", "O.sec"), sep = ":", remove = FALSE, extra = "merge")

regatta.master.times <- regatta.master.times %>%
  mutate(O.min = as.numeric(O.min), O.sec = as.numeric(O.sec))

regatta.master.times <- regatta.master.times %>%
  mutate(total.sec = O.min*60 + O.sec)

## Which event was most competative? 

competitiveness <- regatta.master.times %>% 
  group_by(event.y) %>% 
  summarise(perc.50 = median(total.sec), avg = mean(total.sec), stnd.dev = sd(total.sec)) %>% 
  mutate(cv = stnd.dev/avg)


## what is the "competative" split? 

competitiveness <- competitiveness %>%
  mutate(m.per.sec = 4800/perc.50)

competitiveness <- competitiveness %>%
  mutate(med.split = 500/m.per.sec)


## repeat for entire regatta, including singles to determine most competative race

results.times <- results %>% 
  separate(Raw.Time, into = c("O.min", "O.sec"), sep = 2, remove = FALSE, extra = "merge")
  
results.times <- results.times %>%
  mutate(O.min = as.numeric(O.min), O.sec = as.numeric(O.sec))

results.times <- results.times %>%
  mutate(total.sec = O.min*60 + O.sec)
results.times <- results.times %>% 
  filter(Status == "Finished")

competitiveness.all <- results.times %>% 
  group_by(Event.Id, Event.Description) %>% 
  summarise(perc.50 = median(total.sec), avg = mean(total.sec), stnd.dev = sd(total.sec)) %>% 
  mutate(cv = stnd.dev/avg) %>% 
  arrange(cv)

competitiveness.all <- competitiveness.all %>%
  mutate(m.per.sec = 4800/perc.50)

competitiveness.all <- competitiveness.all %>%
  mutate(med.split = 500/m.per.sec)

competitiveness.all <- competitiveness.all %>% 
  mutate(med.500.split = ifelse(med.split<120, paste("1", round(med.split-60,1), sep = ":"), 
                                  ifelse(med.split>=120 & med.split<130, 
                                         paste("2:0", round(med.split-120,1), sep = ""), 
                                         paste("2", round(med.split-120,1), sep = ":"))))

competitiveness.all <- competitiveness.all %>% 
  ungroup() %>% 
  arrange(cv)
```
```{r, echo=FALSE}

comp.by.event <- plot_ly(  data = competitiveness.all,
  x = Event.Description,
  y = cv,
  text = paste(Event.Description, "<br>", "cv: ", round(cv,4), "500m:", med.500.split),
  title = "Event Competitiveness",
  type = "bar") %>% 
  layout( title = "Event Competitiveness", yaxis = list(title = "CV of Raw Times"))
comp.by.event
```


## Competitiveness for Men's events

```{r, echo=FALSE, warning = FALSE}
race.sex.cox <- select(races, event_id, sex, cox)
competitiveness.all.sex <- left_join(competitiveness.all,
                                     race.sex.cox,
                                     by = c("Event.Id" = "event_id"))

men.competitiveness <- filter(competitiveness.all.sex, 
                              sex == "m")

cme.plot <- ggplot(data = men.competitiveness, 
                   aes(x=factor(Event.Id, levels=unique(Event.Id)), 
                       y = cv))+
  geom_bar(stat = "identity", 
           aes(text = paste(Event.Description, "<br>", 
                        "500m:", med.500.split), 
           fill = cox))+
  ggtitle("Men's Events by Competitiveness")+
  xlab("Event ID")+
  ylab("CV of Times")+
  geom_hline(yintercept = median(men.competitiveness$cv))
ggplotly(cme.plot)
```


## Competitiveness for Women's events

```{r, echo = FALSE, warning = FALSE}
women.competitiveness <- filter(competitiveness.all.sex, 
                              sex == "w")
women.competitiveness <- women.competitiveness %>% 
  ungroup %>% 
  arrange(cv)

cwe.plot <- ggplot(data = women.competitiveness, 
                   aes(x=factor(Event.Id, levels=unique(Event.Id)), 
                       y = cv))+
  geom_bar(stat = "identity", 
           aes(text = paste(Event.Description, "<br>", 
                        "500m:", med.500.split), 
           fill = cox))+
  ggtitle("Women's Events by Competitiveness")+
  xlab("Event ID")+
  ylab("CV of Raw Times")+
   geom_hline(yintercept = median(women.competitiveness$cv))
ggplotly(cwe.plot)
```


## 2. Penalties
```{r, echo = FALSE, fig.height=4}
penalty.by.race <- results %>% 
  mutate(penalty.y.n = ifelse(Penalty > 0, 1, 0)) 

penalty.by.race <- penalty.by.race %>% 
  group_by(Event.Id, Event.Description) %>% 
  summarise(num.pen = sum(penalty.y.n)) 

races <- races %>% 
  mutate(event_id = as.character(event_id))

penalty.by.race <- left_join(penalty.by.race, races, by  = c("Event.Id" = "event_id")) 

penalty.by.race <- penalty.by.race %>% 
  select(Event.Id, Event.Description, num.pen, entries, cox)

penalty.by.race <- penalty.by.race %>% 
  mutate(perc.pen = as.numeric(round(num.pen/entries, 2)))

penalty.by.race <- penalty.by.race %>%
  ungroup() %>% 
  arrange(desc(perc.pen))

pen.by.race.plot <- ggplot(data = penalty.by.race, aes(x=factor(Event.Id, levels=unique(Event.Id)), y = perc.pen))+
  geom_bar(stat = "identity", aes(fill = cox, 
                                  text = paste(
                                    Event.Description, "<br>",
                                    "Entries: ", entries)))+
  ggtitle("Penalties by Race")+
  xlab("Event Number")+
  ylab("Percentage of Boats Receiving a Penalty")
ggplotly(pen.by.race.plot)
```


## The Course
```{r, echo = FALSE}
start.finish.map <- leaflet() %>% 
  addTiles() %>% 
  addPopups(-71.10782, 42.35294, "Start, BU Boathouse") %>% 
  addPopups(-71.13654, 42.36736, "Finish")
start.finish.map
```


## The Course 

```{r, echo = FALSE}
markers.map <- leaflet(data = markers) %>% 
  addTiles() %>%
  addMarkers(~long, ~lat, popup = ~name)
markers.map
```


## Penalty Types

Coxed boats only recieved three types of penalties

1. Buoy Violations
2. Arch Violations 
3. Interference/ Failure to yield

## Where do Penalties Happen? 

```{r, echo = FALSE, warning=FALSE, fig.height=4}
penalty.by.loc <- regatta.full %>% 
  filter(penalty!="")

penalty.by.loc <- penalty.by.loc %>% 
  separate(penalty_desc,into = c("p.1", "p.2", "p.3"), sep = ",", remove = FALSE, extra = "merge")

penalty.by.loc <- penalty.by.loc %>% 
  select(event_id, club, p.1,p.2,p.3)

penalty.by.loc <- penalty.by.loc %>% 
  mutate(p.ext = ifelse(p.1 == "8-BG(2)", "8-BG", NA))

penalty.by.loc <- penalty.by.loc %>% 
  gather(key = p.num, value = penalty, p.1:p.ext)

penalty.by.loc <- penalty.by.loc %>% 
  filter(!is.na(penalty))

penalty.loc <- penalty.by.loc %>% 
  separate(penalty, into = c("p.loc","p.type"), sep = "-", remove = FALSE, extra = "merge")

penalty.loc <- penalty.loc %>% 
  mutate(loc = as.numeric(ifelse(p.loc=="7A",7,p.loc)))

penalty.loc <- penalty.loc %>% 
  group_by(loc) %>% 
  tally()
penalty.markers <- left_join(penalty.markers, penalty.loc, by = c("group"="loc"))
penalty.markers <- penalty.markers %>% 
  mutate(n = ifelse(is.na(n), 0, n))

penalty.map <- ggplot(penalty.markers, aes(x=long, y=lat, group=group)) +
  labs(x="longitude", y="latitude", title="Penalties by Location for Coxed Boats") +
  coord_map() +
  geom_polygon(aes(fill=n, text = paste(name))) +
  geom_path()
ggplotly(penalty.map)
```


## 3. Houw Should I Tell my Rowers to Row the Course?

Concepts to consider: 

1. negative splitting
2. passing
3. bottlenecks

## How do most boats row the course?

```{r, results = "hide", warning = FALSE}
split.eval <- regatta.master.times %>% 
  separate(split1, into = c("s1.min", "s1.sec"), sep = ":", remove = FALSE, extra = "merge")

split.eval <- split.eval %>%
  mutate(s1.min = as.numeric(s1.min), s1.sec = as.numeric(s1.sec))

split.eval <- split.eval %>%
  mutate(s1.total.sec = s1.min*60 + s1.sec)

## repeat for the rest of the splits
## Weld, split 2
split.eval <- split.eval %>% 
  separate(split2, into = c("s2.min", "s2.sec"), sep = ":", remove = FALSE, extra = "merge")

split.eval <- split.eval %>%
  mutate(s2.min = as.numeric(s2.min), s2.sec = as.numeric(s2.sec))

split.eval <- split.eval %>%
  mutate(s2.total.sec = s2.min*60 + s2.sec)
## Cambridge, Split 3
split.eval <- split.eval %>% 
  separate(split3, into = c("s3.min", "s3.sec"), sep = ":", remove = FALSE, extra = "merge")

split.eval <- split.eval %>%
  mutate(s3.min = as.numeric(s3.min), s3.sec = as.numeric(s3.sec))

split.eval <- split.eval %>%
  mutate(s3.total.sec = s3.min*60 + s3.sec)
## Finish, split 4
split.eval <- split.eval %>% 
  separate(split4, into = c("s4.min", "s4.sec"), sep = ":", remove = FALSE, extra = "merge")

split.eval <- split.eval %>%
  mutate(s4.min = as.numeric(s4.min), s4.sec = as.numeric(s4.sec))

split.eval <- split.eval %>%
  mutate(s4.total.sec = s4.min*60 + s4.sec)
## Group by race, calculate cv, then take average cv for each split

split.eval.cv <- split.eval %>% 
  group_by(event_id, event.x) %>% 
  summarise(s1.avg = mean(s1.total.sec), s1.stnd.dev = sd(s1.total.sec),
            s2.avg = mean(s2.total.sec), s2.stnd.dev = sd(s2.total.sec),
            s3.avg = mean(s3.total.sec), s3.stnd.dev = sd(s3.total.sec),
            s4.avg = mean(s4.total.sec), s4.stnd.dev = sd(s4.total.sec)) %>% 
  mutate(s1.cv = s1.stnd.dev/s1.avg,
         s2.cv = s2.stnd.dev/s2.avg,
         s3.cv = s3.stnd.dev/s3.avg,
         s4.cv = s4.stnd.dev/s4.avg)
## not condensing into one row
sum.split.eval.cv <- split.eval.cv %>% 
  summarise(s1.med.cv = median(s1.cv), s1.avg.cv = mean(s1.cv),
            s2.med.cv = median(s2.cv), s2.avg.cv = mean(s2.cv),
            s3.med.cv = median(s3.cv), s3.avg.cv = mean(s3.cv),
            s4.med.cv = median(s4.cv), s4.avg.cv = mean(s4.cv))
sum.split.eval.cv.test <- split.eval.cv %>% 
  summarise_each(funs(median, mean), s1.cv, s2.cv, s3.cv, s4.cv)
## try gather then group
sum.split.eval.cv.med <- split.eval.cv %>% 
  select(event_id, s1.cv, s2.cv, s3.cv, s4.cv) %>% 
  gather(key = split, value = cv, s1.cv:s4.cv) %>% 
  group_by(split) %>% 
  summarize(med.cv = median(cv))
sum.split.eval.cv.avg <- split.eval.cv %>% 
  select(event_id, s1.cv, s2.cv, s3.cv, s4.cv) %>% 
  gather(key = split, value = cv, s1.cv:s4.cv) %>% 
  group_by(split) %>% 
  summarize(avg.cv = mean(cv))
sum.split.eval.cv.both <- left_join(sum.split.eval.cv.avg, sum.split.eval.cv.med,
                                    by = "split")
## merge with polygons
sum.split.eval.cv.both <- sum.split.eval.cv.both %>% 
  mutate(group = ifelse(split == "s1.cv", 1, 
                        ifelse(split == "s2.cv",2,
                               ifelse(split == "s3.cv",3,
                                      ifelse(split =="s4.cv", 4,0)))))
split.polygons <- left_join(split.polygons, sum.split.eval.cv.both, by = "group")

```
```{r, echo=FALSE}
split.map.med <- ggplot(split.polygons, aes(x=long, y=lat, group=group)) +
  labs(x="longitude", y="latitude", title="Median CV of splits") +
  coord_map() +
  geom_polygon(aes(fill = med.cv, text = paste(split.x, "<br>",
                                               "Median: ", round(med.cv,4), "<br>",
                                               "Average: ", round(avg.cv, 4))))  +
  geom_path()
ggplotly(split.map.med)
```

## The Big Turn

Starboards, up Pressure

1. Hawthorn St. to Belmont Hill Boat House
2. A good line is key
3. Lots of variation in times
4. Ability to pass boats is limited

## How did Midd do? 

The Men

```{r, echo = FALSE}

midd.men <- regatta.master.times %>% 
  filter(event_id == "26")

midd.men.hist <- ggplot(data = midd.men, aes(x=total.sec))+
  geom_histogram(binwidth = 10, aes(text = 
                                      paste("Club: ", club,
                                            "<br>", "Place: ",
                                            place,"<br>",
                                            "Time: ",                                                time_official)))+
  ggtitle("Event 26: Men's Collegiate 4+")+
  geom_vline(xintercept = median(midd.men$total.sec), 
             aes(text = paste("Median Boat")))+
  geom_vline(xintercept = 1091.687, color = "blue")+
  geom_vline(xintercept = 1122.187, color = "#00BCD8")
ggplotly(midd.men.hist)
```


## How did Midd do? 

```{r, echo = FALSE, warning = FALSE, fig.height=4}
midd.men.split <- split.eval %>% 
  filter(event_id == "26") %>% 
  select(place, club, s1.total.sec, 
         s2.total.sec, s3.total.sec, s4.total.sec) %>% 
  gather(key = split, value = total.sec, s1.total.sec:s4.total.sec)

midd.men.split <- midd.men.split %>% 
  mutate(split = ifelse(split=="s1.total.sec", "1: Riverside",
                        ifelse(split=="s2.total.sec", "2: Weld",
                               ifelse(split=="s3.total.sec",
                                      "3: Cambridge",
             ifelse(split=="s4.total.sec","4: Finish",0)))))

midd.men.split.wrap <- 
  ggplot(data = midd.men.split, aes(x=total.sec, fill = place))+
    geom_histogram(aes(text = paste(place,": ", club)))+
    scale_fill_gradient()+
    facet_wrap(~split, scales = "free")+
    ggtitle("Event 26: Men's College 4+")
ggplotly(midd.men.split.wrap)
  
```


## How did Midd do?

The Women

```{r, echo = FALSE}
midd.women <- regatta.master.times %>% 
  filter(event_id == "51")

midd.women.hist <- ggplot(data = midd.women, aes(x=total.sec))+
  geom_histogram(binwidth = 10, aes(text = 
                                      paste("Club: ", club,
                                            "<br>", 
                                            "Place: ", place,
                                            "<br>", "Time: ",                                               time_official)))+
  ggtitle("Event 51: Women's Collegiate 8+")+
  geom_vline(xintercept = median(midd.women$total.sec), 
             aes(text = paste("Median Boat")))+
  geom_vline(xintercept = 1114.910, color = "blue")
ggplotly(midd.women.hist)
```


## How did Midd Do? 

```{r, echo = FALSE, warning=FALSE,fig.height=4}

midd.women.split <- split.eval %>% 
  filter(event_id == "51") %>% 
  select(place, club, s1.total.sec, 
         s2.total.sec, s3.total.sec, s4.total.sec) %>% 
  gather(key = split, value = total.sec, s1.total.sec:s4.total.sec)

midd.women.split <- midd.women.split %>% 
  mutate(split = ifelse(split=="s1.total.sec", "1: Riverside",
                        ifelse(split=="s2.total.sec", "2: Weld",
                               ifelse(split=="s3.total.sec",
                                      "3: Cambridge",
             ifelse(split=="s4.total.sec","4: Finish",0)))))


midd.women.split.wrap <- 
  ggplot(data = midd.women.split, aes(x=total.sec, fill = place))+
    geom_histogram(aes(text = paste(place,": ", club)))+
    scale_fill_gradient()+
    facet_wrap(~split, scales = "free")+
    ggtitle("Event 51: Women's College 8+")
ggplotly(midd.women.split.wrap)
```