---
output: 
  html_document: 
    keep_md: yes
author: "Delaney Moran"
---
# Project Synopsis

## Differential Distributions of Starbucks and Dunkin' Donuts in Eastern MA

#####Delaney Moran 

#####May 2016

### Abstract: 

  The present analysis explores the comparative distributions of Dunkin’ Donuts and Starbucks in eastern Massachusetts. It pulls income and demographic data from [Social Explorer](http://old.socialexplorer.com/pub/reportdata/HtmlResults.aspx?reportid=R11178485) and location data from the Google Places API radar searches. The MA shapefile was loaded [here](https://catalog.data.gov/dataset/tiger-line-shapefile-2015-state-massachusetts-current-census-tract-state-based-shapefile).

  Mapping both franchises in the six counties around Boston revealed that there are many more Dunkin’ Donuts than Starbucks locations in eastern Massachusetts. Additionally, the two franchises have different geographic distributions and appear to cater to different clientele, both spatially and economically. Starbucks are more densely distributed in cities and located in areas with higher median incomes. Dunkin' Donuts are more densely distributed in the suburbs, and located in areas with lower median incomes.



```{r, echo=FALSE, warning = FALSE, error = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggmap))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(rgdal))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(jsonlite))
suppressPackageStartupMessages(library(RCurl))
suppressPackageStartupMessages(library(maptools))
suppressPackageStartupMessages(library(rgeos))
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(rvest))
suppressPackageStartupMessages(library(spatialEco))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(knitr))

Eastern <- c(5, 9, 17, 21, 23, 25)

#downloaded from social explorer ACS2014
MAincomedata <-read.csv("./data/MAincomedata.csv")

#filter to only tracts in counties in eastern MA and select relevant columns
MAincomedata <- MAincomedata %>%
  filter(Geo_STATE == 25) %>%
  filter(Geo_COUNTY %in% Eastern) %>%
  select(Geo_NAME, Geo_QName, Geo_STATE, Geo_COUNTY, Geo_FIPS, Geo_TRACT,
         SE_T001_001, SE_T002_002, SE_T057_001,
         SE_T056_001,
         SE_T056_002,
         SE_T056_003,
         SE_T056_004,
         SE_T056_005,
         SE_T056_006,
         SE_T056_007,
         SE_T056_008,
         SE_T056_009,
         SE_T056_010,
         SE_T056_011,
         SE_T056_012,
         SE_T056_013,
         SE_T056_014,
         SE_T056_015,
         SE_T056_016,
         SE_T056_017) %>%
  rename(total_population = SE_T001_001,
         pop_density = SE_T002_002,
         med_inc = SE_T057_001) %>%
  filter(!is.na(med_inc)) %>%
  mutate(Geo_FIPS = as.factor(Geo_FIPS)) %>%
  mutate(County = ifelse(Geo_COUNTY == 5, "Bristol",
                  ifelse(Geo_COUNTY == 9, "Essex",
                  ifelse(Geo_COUNTY == 17, "Middlesex",
                  ifelse(Geo_COUNTY == 21, "Norfolk",
                  ifelse(Geo_COUNTY == 23, "Plymouth",
                  ifelse(Geo_COUNTY == 25, "Suffolk", NA)))))))

#Load MA shapefiles
#https://catalog.data.gov/dataset/tiger-line-shapefile-2015-state-massachusetts-current-census-tract-state-based-shapefile
shapefile_path <- "./data/tl_2015_25_tract/"
map_obj <- readOGR(dsn = shapefile_path, layer = "tl_2015_25_tract")

# Convert Shapefile to ggplot Data Frame
map_obj$id <- rownames(map_obj@data)
map_points <- fortify(map_obj, region="id")
map_df <- inner_join(map_points, map_obj@data, by="id")

MAmap <- ggplot() +
  geom_polygon(aes(x = long, y = lat, group=group), data=map_df) +
  labs(x="Longitude", y="Latitude") +
  geom_path(size = 1.5) +
  coord_map()

map_census <- inner_join(map_df, MAincomedata, by = c("GEOID" = "Geo_FIPS"))

sbloc <- read.csv("./data/sbloc.csv")
ddloc <- read.csv("./data/ddloc.csv")

#SB locations
sb_coord <- sbloc %>%
  select(long, lat)
sb_data <- sbloc %>%
  select(place_id)
CRS <- "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0 "
sbloc_sp <- SpatialPointsDataFrame(sb_coord, sb_data, proj4string = CRS(CRS))

map_with_sb <- point.in.poly(sbloc_sp, map_obj)
map_with_sb_id <- merge(sbloc_sp, map_with_sb, by = "place_id")
map_with_sb_id <- sp.na.omit(map_with_sb_id)


#DD locations
dd_coord <- ddloc %>%
  select(long, lat)
dd_data <- ddloc %>%
  select(place_id)
CRS <- "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0 "
ddloc_sp <- SpatialPointsDataFrame(dd_coord, dd_data, proj4string = CRS(CRS))

map_with_dd <- point.in.poly(ddloc_sp, map_obj)
map_with_dd_id <- merge(ddloc_sp, map_with_dd, by = "place_id")
map_with_dd_id <- sp.na.omit(map_with_dd_id)


#Count DDs and SBs by census tract
DDlist <- as.data.frame(map_with_dd_id) %>%
  select(place_id, lat, long, GEOID) %>%
  rename(dd_place_id = place_id,
         dd_lat = lat,
         dd_long = long) %>%
  mutate(GEOID = as.factor(as.character(GEOID)))

SBlist <- as.data.frame(map_with_sb_id) %>%
  select(place_id, lat, long, GEOID) %>%
  rename(sb_place_id = place_id,
         sb_lat = lat,
         sb_long = long) %>%
  mutate(GEOID = as.factor(as.character(GEOID)))

tracts_w_dd <- as.data.frame(map_obj) %>%
  select(GEOID, NAMELSAD) %>%
  left_join(DDlist, by = "GEOID") %>%
  mutate(dd = ifelse(is.na(dd_place_id), 0, 1)) %>%
  group_by(GEOID) %>%
  summarise(numDD = sum(dd))


tracts_w_sb <- as.data.frame(map_obj) %>%
  select(GEOID, NAMELSAD) %>%
  left_join(SBlist, by = "GEOID") %>%
  mutate(sb = ifelse(is.na(sb_place_id), 0, 1)) %>%
  group_by(GEOID) %>%
  summarise(numSB = sum(sb))

tracts_w_SBDD <- left_join(tracts_w_dd, tracts_w_sb, by = "GEOID")

MAincomedata2 <- left_join(MAincomedata, tracts_w_SBDD, by = c ("Geo_FIPS" = "GEOID")) %>% 
    mutate(sb_1000 = numSB * 1000 / total_population,
         sb_1000 = ifelse(sb_1000 == Inf, NA, sb_1000),
         sb_1000 = ifelse(total_population < 1000, NA, sb_1000),
         dd_1000 = numDD * 1000 / total_population,
         dd_1000 = ifelse(dd_1000 == Inf, NA, dd_1000),
         dd_1000 = ifelse(total_population < 1000, NA, dd_1000))

totals <- MAincomedata2 %>%
  summarise(Starbucks_Count = sum(numSB),
            DunkinDonuts_Count = sum(numDD))

#Calculate franchises per 1000
#There is one census tract with a very small population (480), but 5 Dunkin' Donuts
#and 6 Starbucks...this seems like an anomaly. I am filtering out all tracts that
#have fewer than 1000 people in them
map_census2 <- left_join(map_census, tracts_w_SBDD, by = "GEOID") %>%
  mutate(sb_1000 = numSB * 1000 / total_population,
         sb_1000 = ifelse(sb_1000 == Inf, NA, sb_1000),
         sb_1000 = ifelse(total_population < 1000, NA, sb_1000),
         dd_1000 = numDD * 1000 / total_population,
         dd_1000 = ifelse(dd_1000 == Inf, NA, dd_1000),
         dd_1000 = ifelse(total_population < 1000, NA, dd_1000))

MAmap +
  geom_point(data = ddloc,
             aes(x = long, y = lat),
             color = 'orange',
             size = 1.5) +
  geom_point(data = sbloc,
             aes(x = long, y = lat),
             color = 'green',
             size = 1.5) +
  coord_map() +
  theme_bw() +
  labs(title = "Locations of Starbucks and Dunkin' Donuts in Eastern MA")

```

__Figure 1.__ Orange dots represent Dunkin Donuts locations and green dots represent Starbucks locations in the greater Boston area.


```{r, echo= FALSE, warning=FALSE}
inc_data <- MAincomedata2 %>% 
  select(Geo_FIPS, dd_1000, sb_1000, med_inc) %>% 
  filter(!is.na(dd_1000), !is.na(sb_1000))
inc_data <- gather(inc_data, "Franchise", "Density", ends_with("1000")) %>% 
  mutate(Franchise = ifelse(Franchise == "dd_1000", 
                            "Dunkin_Donuts", "Starbucks"))

dd_data2 <- filter(inc_data, Franchise == "Dunkin_Donuts") 
dd_inc <- lm(Density ~ med_inc, data = dd_data2)
ddinccoefs <-coefficients(dd_inc)


sb_data2 <- filter(inc_data, Franchise == "Starbucks")
sb_inc <- lm(Density ~ med_inc, data = sb_data2)
sbinccoefs <- coefficients(sb_inc)



```


```{r, echo= FALSE, warning=FALSE}

ggplot(sb_data2, aes(x = med_inc, y = Density)) +
  geom_point() +
  geom_abline(slope = sbinccoefs[2], intercept = sbinccoefs[1], col = "darkgreen", size = 1.5) +
  xlab("Median Income") +
  ylab("Franchise Density")+
  ggtitle("Starbucks Density by Median Income")

ggplot(dd_data2, aes(x = med_inc, y = Density)) +
  geom_point() +
  geom_abline(slope = ddinccoefs[2], intercept = ddinccoefs[1], col = "darkorange", size =1.5) +
  xlab("Median Income") +
  ylab("Franchise Density")+
  ggtitle("Dunkin' Donuts Density by Median Income")


```


__Figure 2.__ A linear regression revealed that median income does significantly predict Starbucks densities, such that census tracts with higher median incomes have more Starbucks. This effect was subtle (a very small slope), but nonetheless significant (p = 0.0017). For Dunkin' Donuts, the relationship was only marginally significant (p = 0.0660), and equally subtle, but, interestingly, of opposite magnitude! In census tracts with higher median incomes, there were fewer Dunkin' Donuts.
