---
output: 
  html_document: 
    keep_md: yes
---


# Project Details

* Name: Carter Merenstein
* Project title: Lyme Disease and the Tick Microbiome
* Abstract: Lyme disease is the most common vector borne disease in the United States, with over 30,000 cases reported each year. It is caused by the bacteria *Borrelia burgdorferi*, which is carried by ticks and infects humans after a tick bite. In this project I looked at the bacterial microbiome of 92 ticks in order to try to identify bacteria that might reduce *Borrelia* prevalence. In the process I did an exploratory data analysis and identified differences in the microbiomes of ticks by sex and location. I was unable to identify any bacteria that were significantly negatively correlated with *Borrelia* prevalence, though I did find that bacteria in the family *Oxalobacteraceae* were positively correlated with *Borrelia* abundance.

Below are stacked bar graphs which represent the microbiomes from all 92 ticks. Colors represent different types of bacteria, and each virtical bar is a tick. This is an easy graphical way to begin to understand differences in tick microbiomes.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}

## I know this is a ton of code for a readme, but it's necessary for just one graph.


suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(randomForest))
suppressPackageStartupMessages(library(stringr))


#this global is used for random forests later
# at 50000 the models get stable, but 500 is usefull for kntting fast.
NUM_TREES = 500

# This vector of colors is used for the stacked bar graphs.
c25 <- c("#6A3D9A", # purple
         "dodgerblue2","#E31A1C", # red
         "green4",
         "#FF7F00", # orange
         "Grey","gold1",
         "steelblue4","#FB9A99", # lt pink
         "palegreen2",
         "#CAB2D6", # lt purple
         "maroon","#FDBF6F", # lt orange
         "khaki2",
         "orchid1","deeppink1","blue1",
         "darkturquoise","green1","yellow4","yellow3","skyblue2",
         "darkorange4","brown", "gray70")

# These two other vectors are used for certain subseted stacked bar graphs
# They have colors removed to ensure that color-coding is consistent
c24 <- c25
c24[24] = "gray70" ## for females
cN <- c25[-13][-20] ## for negatives

# Data are in long format; each row is a tick, each column an OTU
ticks <- read.csv("data/ALL_Ticks.csv")
ticks_long <- ticks #this gets used later
ticks_long.orig <- ticks_long
ticks <- ticks %>% gather(OTU, Count, 2:ncol(ticks))
ticks.orig <- ticks
# Turn count into percent
stacks <- group_by(ticks, Tick, OTU) %>%  summarize(count = sum(Count))  %>% 
  filter(count > 0) %>% 
  inner_join(group_by(ticks, Tick) %>% summarize( Total = sum(Count))) %>% 
  mutate(Percent = ((count/Total) * 100) )

#taxonomy file has full taxonomic information of each OTU
taxonomy <- read.csv("data/OTU_Taxonomy.txt", sep = "\t") %>% tbl_df() %>% 
  mutate(Genus = gsub('[()0-9 ]', '', Genus)) %>% 
  mutate(Family = gsub('[()0-9 ]', '', Family)) %>% 
  mutate(Order = gsub('[()0-9 ]', '', Order))

# get OTU in same format
taxonomy <- mutate(taxonomy, OTU = gsub('[A-Z,a-z.]', '', OTU))
stacks <- mutate(stacks, OTU = gsub('[A-Z,a-z.]', '', OTU))

# the Taxon variable tells us the lowest identified taxonomical information
stacks <- inner_join(stacks, taxonomy, by="OTU") %>% 
  mutate(Taxon = ifelse(nchar(Genus) < 4, ifelse(nchar(Family) < 4, ifelse(nchar(Order) < 4, as.character(taxonomy), Order), Family), Genus)) %>% 
  mutate(OTU_taxon = paste(OTU, Taxon, sep = "; "))

# this is a totally janky section that is necessary because of some unidentified bug. Somewhere along the line the percents got messed up, this fizes them
stacks <- group_by(stacks, Tick) %>% summarize(Tot = sum(count)) %>% 
    inner_join(stacks, by = "Tick") %>% 
    mutate(p2 = count/Tot) %>% 
    mutate(Percent = p2 * 100)

# for stacked bar chart, we want only the top 24 most common, everything else gets put into "other"
most_common_df <- stacks %>% group_by(OTU_taxon) %>% 
    summarize(Total = sum(Percent)) %>%  top_n(24)
stacks_collapsed <- mutate(stacks, OTU_Collapsed = ifelse(OTU_taxon %in% most_common_df$OTU_taxon, OTU_taxon, "Other")) %>% 
  group_by(Tick, OTU_Collapsed)  %>% summarize(Percent = sum(Percent))

# the names tell what state they're from
# this was initially necessary because the metadata wasn't availible
# Now I could use the metadata file mf.txt, but this code works a well
NC_affinis = c("BLA", "HAL", "JON", "WAS")
NC_scapularis = c("FT", "Ixs")
SC = c("DRGF", "DRGM", "DRHF")

stacks_collapsed <- mutate(stacks_collapsed, region = gsub('[0-9.]+', '', Tick))

stacks_collapsed <- mutate(stacks_collapsed, region = ifelse(region %in% NC_affinis, "NC_a", region)) %>% 
  mutate(region = ifelse(region %in% NC_scapularis, "NC_s", region)) %>% 
  mutate(region = ifelse(region == "RJB", "VA", region)) %>% 
  mutate(region = ifelse(region %in% SC, "SC", region))

stacks_collapsed$region = as.factor(stacks_collapsed$region)

#Without further ado: stacked bar chart
stacked_bar_chart <- ggplot(stacks_collapsed, aes(x = Tick, y = Percent, fill = OTU_Collapsed)) +
    geom_bar(stat = "identity", color = "black") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(legend.position="bottom", text=element_text(size=14), legend.key.size=unit(.25, "cm")) +
    scale_fill_manual(values = c25) +
      labs(fill = "OTU; Taxon") + 
    guides(fill = guide_legend(ncol = 3)) +
    theme(legend.background = element_rect(color = "Black", size = .5, linetype = "solid"))
```


```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}

metadata <- read.csv("data/mf.txt", sep = '\t') %>% 
  rename(Tick = X.SampleID)
sex <- metadata %>%  select(`Tick`, `Sex`)

stacks_collapsed <- inner_join(stacks_collapsed, sex)

# it just looks better when they're two separate graphs, not faceted
f_collapsed = filter(stacks_collapsed, Sex == "F")
m_collapsed = filter(stacks_collapsed, Sex == "M")

stacked_bar_f <-ggplot(f_collapsed, aes(x = Tick, y = Percent, fill = OTU_Collapsed)) +
  geom_bar(stat = "identity", color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position="none" ,text=element_text(size=14)) +
  scale_fill_manual(values = c24) + 
  facet_grid(.~ region, scale="free", space="free") +
  labs(x = "Female Ticks")

stacked_bar_f
```

```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
stacked_bar_m <-ggplot(m_collapsed, aes(x = Tick, y = Percent, fill = OTU_Collapsed)) +
  geom_bar(stat = "identity", color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position="bottom", text=element_text(size=14), legend.key.size=unit(.25, "cm")) +
  scale_fill_manual(values = c25) +
  facet_grid(.~ region, scale="free", space="free") +
  labs(x = "Male Ticks") +
  labs(fill = "OTU; Taxon") + 
    guides(fill = guide_legend(ncol = 3)) +
    theme(legend.background = element_rect(color = "Black", size = .5, linetype = "solid"))
stacked_bar_m

```
